---
title: "Data Science project - Gen AI misinformation Detection Data"
output: html_notebook
---

# STEP 1 - INITIAL SETUP

```{r}
# Install all necessary packages
# install.packages(c("tidyverse", "readr", "ggplot2","maps", "tidymodels",  "purrr", "patchwork", "scales", "magrittr", "pheatmap", "treemapify", "knitr","recipes", "yardstick"))

# Load all necessary packages
library(readr)
library(tidyverse)
library(ggplot2)
# library(scales) # for percentages
library(dplyr)
library(maps) # for geographical representation
library(tidymodels) # for ML
library(pheatmap) # for correlation matrix
library(lubridate)
library(parsnip) # for ML
library(skimr)
library(treemapify)
library(knitr) # for exporting tables
library(yardstick)
library(recipes)


# load the data set 
generative_ai_misinformation_dataset <- read_csv("~/Desktop/ECON4170 Data Science for Economists /generative_ai_misinformation_dataset.csv")

```

# defining theme colors for a unified interface

```{r}
# define color palettes
misinfo_colors <- c("0" = "#2ECC71", "1" = "#E74C3C")

# using the official color codes for a more intuitive visualization
platform_colors <- c("Facebook" = "#3b5998", "Twitter" = "#1DA1F2","Reddit" = "#FF4500", "Telegram" = "#0088cc")

```

# STEP 2 - DATA CLEANING / TIDY DATA

```{r}
df <- generative_ai_misinformation_dataset

view(df) #open the entire dataset in a separate window
glimpse(df) # using glimpse to check the if R managed to decode variable types right 
tibble(df)
summary(df)
skimr::skim(df) # investigating missing values
skim(df)

# removing unnecessary variables
df <- df %>%
  select(-c(text, timezone, post_id, id, author_id))

# converting categorical data with as.factor
df <- df %>%
  mutate(
    # convert temporal data
    timestamp = as.POSIXct(timestamp),
    date = as.Date(date), # or: as_date
    # convert other categorical data to factorial # obs across(where(is.character), as.factor) cannot be used here
    is_misinformation = as.factor(is_misinformation),
    platform = as.factor(platform),
    author_verified = as.factor(author_verified),
    model_signature = as.factor(model_signature),
    factcheck_verdict = as.factor(factcheck_verdict),
    country = as.factor(country),
    city = as.factor(city),
    weekday = as.factor(weekday),
    month = as.factor(month),
    # extracting time components from the timestamp
    hour = hour(timestamp),
    day_of_week = wday(timestamp, label = TRUE),
    month_num = month(timestamp),
    year = year(timestamp)
  )

# after extracting components from the data frame, we no longer need time stamp
df$timestamp <- NULL

# checking if all necessary variables are converted right
glimpse(df)

```

# STEP 3 - DATA WRANGLING

# 3.1 General misinforamtion distribution

```{r}
# version 1 - barplot

misinformation_distribution <- df %>%
  count(is_misinformation) %>%
  mutate(percentage = n / sum(n)) %>%
  ggplot(aes(x = is_misinformation, y = n, fill = is_misinformation)) +
  geom_col() +
  # printing percentages over bars
  geom_text(aes(label = sprintf("%.1f%%", percentage * 100)), 
            # bringing the percentage labels up from the bars (otherwise number merge with the bars)
            vjust = -0.5) +
  # incorporating unified colors
  scale_fill_manual(values = misinfo_colors, labels = c("0" = "Authentic","1" = "Misinformation")) +
  labs(title = "Misinformation Distribution", subtitle = "Across the Dataset", x = "Is Misinformation", y = "Posts") +
  theme_minimal() + 
  # adjusting the style of the titles
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5))

print(misinformation_distribution)

```

```{r}
# version 2 - pie chart  

misinformation_pie <- df %>%
  count(is_misinformation) %>%
  mutate(
    percentage = n / sum(n),
    label = paste0(labels = c("Authentic", "Misinformation")[as.numeric(is_misinformation)+1], "\n", scales::percent(percentage, accuracy = 0.1))) %>%
  ggplot(aes(x = "", y = percentage, fill = is_misinformation)) +
  geom_col(color = "white", width = 1) + # slices
  coord_polar(theta = "y") + # converting to pie
  scale_fill_manual(values = misinfo_colors, labels = c("0" = "Authentic", "1" = "Misinformation")) +
  labs(title = "Misinformation distribution", fill = "Classification") +
  theme_void() +
  # theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5, face = "bold", size = 16)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16)) +
  # labeling slices
  geom_text(
    aes(label = scales::percent(percentage, accuracy = 0.1)),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 5)

misinformation_pie
```

# 3.2 Per platform (significant)

```{r}
misinfo_per_platform <- df %>%
  group_by(platform) %>%
  summarise(
    total_posts   = n(),
    misinfo_posts = sum(is_misinformation == "1"),
    misinfo_index = misinfo_posts / total_posts)

print(misinfo_per_platform)
```

```{r}
posts_per_platform_bar <- df %>%
  count(platform, sort = TRUE) %>% #sort the platform by posts (big->small)
  #visuals
  ggplot(aes(x = reorder(platform, n), y = n, fill = platform)) +
  geom_col() +
  coord_flip() + # flip bars horizontally
  scale_fill_manual(values = platform_colors) +
  labs(title = "Posts by platform", subtitle = "Barplot representation", x = NULL, y = "Number of Posts", fill = "Platform") +
  # theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),) + # adjusting the header
  theme(
    plot.title    = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "bottom")

print(posts_per_platform_bar)

```

```{r}
# obs changed from treemap tp treemapify 

posts_per_platform_treemap <- df %>%
  count(platform, sort = TRUE) %>%       # platform + post counts
  ggplot(aes(area  = n, fill  = platform, label = paste0(platform, "\n", n))) +   # "label" puts text inside tiles
  geom_treemap(color = "white") +        # white borders like gallery
  geom_treemap_text(place = "centre", colour = "white") +
  scale_fill_manual(values = platform_colors) +
  labs(title = "Posts per platform", fill = "Platform") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "bottom")

posts_per_platform_treemap
```

```{r}
  
# misinformation per platform without percentages

misinfo_per_platform_visual <- df %>%
  group_by(platform, is_misinformation) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = platform, y = n, fill = is_misinformation)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = misinfo_colors,
                    labels = c("0" = "Authentic", "1" = "Misinformation")) +
  labs(title = "Misinformation per platform", fill = "Classification", x = "Platform", y = "Percentage") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5), legend.position = "bottom") +
  # labeling bars
  geom_text(
    aes(label = scales::percent(percentage, accuracy = 0.1)),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 5)

# with percentages
misinfo_per_platform_visual <- df %>%
  group_by(platform, is_misinformation) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(platform) %>% 
  # step 1 - compute share per platform
  mutate(percentage = n / sum(n)) %>%
  ggplot(aes(x = platform, y = percentage, fill = is_misinformation)) +
  geom_col(position = "stack") +
  # step 2 - percentage texts
  geom_text(
    aes(label = scales::percent(percentage, accuracy = 0.1)),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 4) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = misinfo_colors, labels = c("0" = "Authentic", "1" = "Misinformation")) +
  labs(title = "Misinformation per platform", x = "Platform", y = "Share", fill = "Classification") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5), legend.position = "bottom")

print(misinfo_per_platform_visual)

```

# 3.3 - Temporal Patterns

# 3.3.1 - by hour

# Per hour - significant

```{r}
misinfo_per_hour <- df %>%
  group_by(hour) %>%
  summarise(misinfo_share = sum(is_misinformation == "1") / n()*100)

# print inside the project
print(misinfo_per_hour)

# print out for report
misinfo_per_hour %>%
  mutate(misinfo_share = round(misinfo_share, 1)) %>%
  knitr::kable(
    col.names = c("Hour of the day", "Misinformation share (%)"),
    caption   = "Misinformation share per hour")

```

```{r}
misinfo_per_hour_visual <- df %>%
  group_by(hour, is_misinformation) %>%
  summarise(n = n(), .groups = "drop") %>%
  
  # creating a visual
  ggplot(aes(x = hour, y = n, fill = is_misinformation)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = misinfo_colors, labels = c("0" = "0 - Authentic", "1" = "1 - Misinformation")) +
  labs(title = "Misinformation share by time of the day", fill = "Classification", x = "Time of the day (hour)", y = "Percentage") +
  theme_minimal() +
  theme(
    plot.title  = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 0, hjust = 1))

print(misinfo_per_hour_visual)

```

```{r}
# circular barplot representation (only misinformation)

# choosing only misinformation and summarising by hourly misinformation share
misinfo_share_hour <- df %>%
  group_by(hour) %>%
  summarise(
    misinfo_posts = sum(is_misinformation == "1"),
    total_posts = n(),
    misinfo_share = misinfo_posts / total_posts)

# visual as suggested by r-graph-gallery.com
misinfo_per_hour_circular <- misinfo_share_hour %>%
  ggplot(aes(x = factor(hour), y = misinfo_share)) +
  geom_col(fill = misinfo_colors["1"], color = "white", width = 1) +
  coord_polar(start = -pi/23) + # rotate to start at top
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Misinformation share by time of the day", x = NULL, y = NULL) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.text.x = element_text(size = 10, face = "bold")
  )

print(misinfo_per_hour_circular)

```

# 3.3.2 - by weekday

```{r}
misinfo_per_weekday <- df %>%
  mutate(weekday = factor(
      weekday,
      levels = c("Monday", "Tuesday", "Wednesday","Thursday", "Friday", "Saturday", "Sunday"),
      ordered = TRUE)) %>%
  group_by(weekday) %>%
  summarise(misinfo_share = sum(is_misinformation == "1") / n()*100)

# print inside the project 
print(misinfo_per_weekday)

# print out for report
misinfo_per_weekday %>%
  mutate(misinfo_share = round(misinfo_share, 1)) %>%
  knitr::kable(
    col.names = c("Day of the week", "Misinformation share (%)"),
    caption   = "Misinformation share weekday")
```

```{r}
# VISUAL
misinfo_per_weekday_visual <- df %>%
  group_by(weekday, is_misinformation) %>%
  summarise(n = n(), .groups = "drop") %>%
  
  # we need to reorder weekdays 
  mutate(
    weekday = factor(
      weekday,
      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"),
      ordered = TRUE)) %>%
  # creating a visual
  ggplot(aes(x = weekday, y = n, fill = is_misinformation)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(
    values = misinfo_colors,
    labels = c("0" = "Authentic", "1" = "Misinformation")) +
  labs(title = "Misinformation by the day of the week", fill = "Classification", x = "Day of the week", y = "Share") +
  theme_minimal() +
  theme(
    plot.title  = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
  

print(misinfo_per_weekday_visual)
```

# 3.3.3 - By month (significant)

```{r}
# months <- unique(df$month)

# obs - we have more observations for some months than for others => look only %-wise on misinformation (aggregate)
misinfo_per_month <- df %>%
  group_by(month) %>%
  summarise(
    total_posts   = n(),
    misinfo_posts = sum(is_misinformation == "1"),
    misinfo_index = misinfo_posts / total_posts)

print(misinfo_per_month)

```

```{r}

# VISUAL
misinfo_per_month_visual <- df %>%
  group_by(month, is_misinformation) %>%
  summarise(n = n(), .groups = "drop") %>%
  # we need to reorder months 
  mutate(
    month = factor(
      month,
      levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"),
      ordered = TRUE)) %>%
  # creating a visual
  ggplot(aes(x = month, y = n, fill = is_misinformation)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(
    values = misinfo_colors,
    labels = c("0" = "Authentic", "1" = "Misinformation")
  ) +
  labs(title = "Misinformation per month", fill = "Classification", x = "Month", y = "Share") +
  theme_minimal() +
  theme(
    plot.title  = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1))

print(misinfo_per_month_visual)

# conclusion - some months there is way more misinformation than other - f.eks. only 41% in June but 65% in May

```

# 3.3.4 - Throughout time - more misinformation nowadays?

```{r}

# we want to create a new step - by month across the whole data frame
df <- df %>%
  mutate(month_year = floor_date(date, "month"))

# grouping
misinfo_with_time <- df %>%
  group_by(month_year) %>%
  summarise(total_posts = n(),
    misinfo_posts = sum(is_misinformation == "1"),
    misinfo_share = misinfo_posts / total_posts
  )

# plotting with ggplot
ggplot(misinfo_with_time, aes(x = month_year)) +
  # grey line - total posts 
  geom_line(aes(y = total_posts, color = "Total posts"), size = 1.2) +
  # red line 
  geom_line(aes(y = misinfo_posts, color = "Misinformation posts"), size = 1.2) +
  labs(title = "Total posts vs Misinformation posts", x = "Month", y = "Number of posts") +
  scale_color_manual(values = c("Total posts" = "darkgrey", "Misinformation posts" = "red")) +
  theme_minimal()+ 
  theme(plot.title  = element_text(size = 16, face = "bold", hjust = 0.5))
```

# looking at the misinformation share over time (step = month)

```{r}

ggplot(misinfo_time, aes(x = month_year, y = misinfo_share)) +
  geom_line(color = "red", size = 1) + # adding misinformation line
  geom_point(color = "red", size = 2) + # point for each month
  scale_y_continuous(labels = scales::percent_format()) + # scale = misinformation share
  # make the step on x-axis more evident
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  labs(title = "Misinformation share over time", x = "Timeline", y = "Misinformation share (%)") +
  theme_minimal() +
  theme(plot.title  = element_text(size = 16, face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Compute misinformation share per month and platform
misinfo_time_platform <- df %>%
  group_by(month_year, platform) %>%
  summarise(misinfo_share = mean(is_misinformation == "1"), .groups = "drop")  # share of misinfo

# plotting
ggplot(misinfo_time_platform, aes(x = month_year, y = misinfo_share, color = platform)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.2, size = 0.8) + # no SE, Locally Estimated Scatterplot Smoothing
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  # make the step on x-axis more evident
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  scale_color_manual(values = platform_colors) +
  labs(title = "Misinformation over time per platform", x = "Time (montly)", y = "Misinformation share", color = "Platform") +
  theme_minimal() +
  theme(
    plot.title  = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1))
```

# excluding Facebook

```{r}
# Compute misinformation share per month and platform
misinfo_time_platform <- df %>%
  filter(platform != "Facebook") %>%
  group_by(month_year, platform) %>%
  summarise(misinfo_share = mean(is_misinformation == "1"), .groups = "drop")  # share of misinfo

# plotting
ggplot(misinfo_time_platform, aes(x = month_year, y = misinfo_share, color = platform)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.2, size = 0.8) + # no SE, Locally Estimated Scatterplot Smoothing
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  # make the step on x-axis more evident
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  scale_color_manual(values = platform_colors) +
  labs(title = "Misinformation over time per platform", x = "Time (montly)", y = "Misinformation share", color = "Platform") +
  theme_minimal() +
  theme(
    plot.title  = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)) + 
  geom_vline(xintercept = as.Date("2024-12-01"), color = "red",linetype = "dashed",linewidth = 1.0) 
```

# 3.4 Per model signature (insignificant)

```{r}
misinfo_per_model_signature <- df %>%
  group_by(model_signature) %>%
  summarise(total_posts   = n(), misinfo_posts = sum(is_misinformation == "1"), misinfo_index = misinfo_posts / total_posts)

print(misinfo_per_model_signature)
```

```{r}

misinfo_per_model_signature_visual <- df %>%
  group_by(model_signature, is_misinformation) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(model_signature) %>% # step 1 in adding % 
  mutate(percentage = n / sum(n)) %>% # step 2 in adding % 
  ggplot(aes(x = model_signature, y = percentage, fill = is_misinformation)) +
  geom_col(position = "stack") +
  # step 3 in adding %
  geom_text(
    aes(label = scales::percent(percentage, accuracy = 0.1)),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 4) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = misinfo_colors,
                    labels = c("0" = "Authentic",
                               "1" = "Misinformation")) +
  labs(title = "Misinformation by model signature", fill = "Classification", x = "Model Signature", y = "Percentage") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5), legend.position = "bottom") 

print(misinfo_per_model_signature_visual)

```

# 3.5 - Geographical spread

```{r}

# loading maps of cities for the world map
data("world.cities")
  
posts_by_city <- df %>%
  group_by(country, city) %>%
  summarise(
    total_posts   = n(),
    misinfo_posts = sum(is_misinformation == "1", na.rm = TRUE),
    # misinfo_index = if_else(total_posts > 0, misinfo_posts / total_posts, 0),
    misinfo_rate = mean(is_misinformation == "1") * 100, .groups = "drop")

# connecting to city coordinates for plotting
city_coordinates <- world.cities %>%
  select(name, country.etc, lat, long)

posts_city_geo <- posts_by_city %>%
  left_join(city_coordinates, by = c("city" = "name", "country" = "country.etc"))

ggplot() +
  borders("world", colour = "grey80", fill = "grey95") +
  geom_point(
    data = posts_city_geo %>% filter(!is.na(lat), !is.na(long)),
    aes(x = long, y = lat, size = total_posts, color = misinfo_rate), alpha = 0.7) + # alpha reduces opacity
  scale_size_continuous(name = "Total posts") +
  # adjusting colors
  scale_color_gradient(low  = "green", high = "red", name = "Misinformation rate (%)") +
  coord_fixed(1.3) + # elongate the picture vertically
  theme_void() + #remove ticks
  # adding labels
  labs(title = "Geographical spread of misinformation", subtitle = "City-level epicenters") +
  # adjusting the style of the titles
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5))

# conclusion: we see a big difference here!
```

```{r}

posts_by_cities_2 <- df %>%
  group_by(city) %>%
  summarise(posts = n(), misinfo_rate = mean(is_misinformation == "1") * 100) %>%
  ggplot(aes(x = reorder(city, misinfo_rate), y = posts)) +
  geom_point(aes(size = posts, color = misinfo_rate)) +
  scale_color_gradient2(low = "green", mid = "yellow", high = "red",midpoint = 50) +
  coord_flip() +
  labs(title = "Post volume and misinformation in cities",x = "City", y = "Post volume") + 
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),)

print(posts_by_cities_2)
```

# STEP 4 - CORRELATIONS IN OTHER FACTORS

# Matrix

```{r}
matrix_variables <- df %>%
select(is_misinformation,
       toxicity_score,
       sentiment_score,
       detected_synthetic_score,
       factcheck_verdict,
       external_factchecks_count, 
       source_domain_reliability,
       embedding_sim_to_facts,
       author_followers,
       author_verified,
       text_length,
       token_count,
       engagement)

# converting categorical factors to numeric in  to use them in the matrix
matrix_variables$is_misinformation <- as.numeric(as.character(matrix_variables$is_misinformation))
matrix_variables$factcheck_verdict <- as.numeric(matrix_variables$factcheck_verdict)
matrix_variables$author_verified <- as.numeric(matrix_variables$author_verified)

corr_matrix <- cor(matrix_variables, use = "complete.obs")

# creating a visualization
pheatmap(corr_matrix,
         display_numbers = TRUE,
         number_format = "%.2f",
         color = colorRampPalette(c("#E74C3C", "white", "#2ECC71"))(200),
         main = "Correlation matrix")
```

# STEP 5 - MACHINE LEARNING

# 5.0 Creating a data frame only with variables we want to use in this step

```{r}
ml_df <- df %>%
  select(
    is_misinformation,
    hour,
    month_num,
    platform,
    city
  )
```

# 5.1 Converting variables from factor to numerical

```{r}
# the features for testing should be numerical and is_misinformation a dummy
ml_df <- ml_df %>%
  mutate(across(where(is.character), as.factor), is_misinformation = as.factor(is_misinformation))
```

# 5.2 Splitting the data

```{r}
set.seed(3170)
split <- initial_split(ml_df, prop = 0.8, strata = is_misinformation)
test <- testing(split)
train <- training(split)

# checking the means of misinformation share across the test and train samples 
# need to convert to numeric to get a number (otherwise - NA)
mean_train <- mean(as.numeric(as.character(train$is_misinformation)))
mean_test  <- mean(as.numeric(as.character(test$is_misinformation)))

mean_train # 0.5363
mean_test # 0.5346
```

# 5.3 Building recipes

```{r}

rec <- recipe(is_misinformation ~., data = train) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())
```

# 5.4 Choosing models

```{r}

# Logistic
log_spec <- logistic_reg() |>
  set_engine("glm") 

# Random forest
rforest_spec <- rand_forest() |>
  set_engine("ranger") |>
  set_mode("classification") # change from: regression

# Gradient boosted tree
xgboost_spec <- boost_tree() |>
  set_engine("xgboost") |>
  set_mode("classification")
```

# 5.3.1 - Training a logistic model

```{r}
# a
log_fit <- log_spec |>
  fit(is_misinformation ~ ., data = train)

# using predict() to predict the outcomes in the test sample 
log_pred <- predict(log_fit, test, type = "prob")

  # get .pred_0 and .pred_1

# b - converting the probabilities to predicted values
log_pred <- log_pred |>
  mutate(pred_class = as.factor(ifelse(.pred_1 >= 0.5, 1, 0)))

# c - creating a confusion matrix 

test_aug <- bind_cols(test, log_pred) |>
  mutate(
    truth = is_misinformation,
    pred_class = factor(pred_class, levels = c(0, 1)))

log_conf <- test_aug |>
  conf_mat(truth = truth, estimate = pred_class)

print(log_conf)

# checking model accuracy using yardstick::accuracy()
# The accuracy-function is basically doing:
# (lm_conf[[1]][1] + lm_conf[[1]][4]) / nrow(monks_test)
# I.e. the (true_neg + true_pos) / n_predictions
lm_acc <- test_aug |> # Save it to compare with xgboost
  accuracy(truth = is_misinformation, estimate = pred_class)

lm_acc
# 0.54 (1st try)
# 0.45 (latest)


```

# Gradient boosting model (XGBoost)

```{r}
# a  - making a new data set
ml_df2 <- ml_df |>
  mutate(class = as.factor(is_misinformation))

set.seed(4170)
df2_split <- initial_split(ml_df2, prop = 0.8, strata = class)
df2_test <- testing(df2_split)
df2_train <- training(df2_split)


# b - specifying and fitting the xgboost model
xgboost_fit <- boost_tree() |>
  set_mode("classification") |>
  fit(is_misinformation ~ ., data = df2_train)


# c - confusion matrix
xgboost_pred <- predict(xgboost_fit, df2_test) |>
  bind_cols(df2_test)

xgboost_pred |>
  conf_mat(truth = is_misinformation, estimate = .pred_class)

# checking accuracy
xgboost_pred |>
  accuracy(truth = is_misinformation, estimate = .pred_class)
```

# Building a workflow

```{r}
wf <- workflow() %>%
  add_recipe(rec)

log_fit <- wf %>% add_model(log_spec) %>% fit(train)
rforest_fit  <- wf %>% add_model(rforest_spec) %>% fit(train)
xgboost_fit <- wf %>% add_model(xgboost_spec) %>% fit(train)


```

# 5.7 Evaluating each model

```{r}
# LOGISTIC
log_pred <- predict(log_fit, test, type = "prob") %>%
  bind_cols(test) %>%
  mutate(
    pred_class = as.factor(ifelse(.pred_1 >= 0.5, 1, 0)))

log_conf <- log_pred %>%
  conf_mat(truth = is_misinformation, estimate = pred_class)

log_acc <- log_pred %>%
  accuracy(truth = is_misinformation, estimate = pred_class)

log_conf
log_acc
# 0.45

# RFOREST
pred_rforest <- predict(rforest_fit, test) %>%
  bind_cols(test)

conf_mat(pred_rforest, truth = is_misinformation, estimate = .pred_class)
accuracy(pred_rforest, truth = is_misinformation, estimate = .pred_class)
# 0.55


#XBOOST
pred_xgboost <- predict(xgboost_fit, test) %>%
  bind_cols(test)

conf_mat(pred_xgboost, truth = is_misinformation, estimate = .pred_class)
accuracy(pred_xgboost, truth = is_misinformation, estimate = .pred_class)
# 0.54



```

# STEP 6 - TUNING (introducing more variables and changing month to categorical)

# 6.0 Choosing new variables for the data frame

```{r}
ml_df2 <- df %>%
  select(
    is_misinformation,
    hour,
    # month_num,
    platform,
    city,
    # new
    month,
    sentiment_score,
    token_count,
    author_followers
  )
```

# 6.1 Converting variables from factor to numerical

```{r}
# the features for testing should be numerical and is_misinformation a dummy
ml_df2 <- ml_df2 %>%
  mutate(
    across(where(is.character), as.factor),
    is_misinformation = as.factor(is_misinformation))
```

# 6.2 Splitting the data

```{r}
set.seed(3170)
split2 <- initial_split(ml_df2, prop = 0.8, strata = is_misinformation)
test2 <- testing(split2)
train2 <- training(split2)

# checking the means of misinformation share across the test and train samples 
# need to convert to numeric to get a number (otherwise - NA)
mean_train2 <- mean(as.numeric(as.character(train2$is_misinformation)))
mean_test2  <- mean(as.numeric(as.character(test2$is_misinformation)))

mean_train2 # 0.5363
mean_test2 # 0.5346

# note: mean stayed the same

```

# 6.3 Building recipes

```{r}
rec2 <- recipe(is_misinformation ~., data = train2) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())
```

# 6.4 Model specification

```{r}

# Logistic
log_spec2 <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")

# Random forest
rforest_spec2 <- rand_forest() |>
  set_engine("ranger") |>
  set_mode("classification") # change from: regression

# Gradient boosted tree
xgboost_spec2 <- boost_tree() |>
  set_engine("xgboost") |>
  set_mode("classification")
```

# 6.5 Building a workflow

```{r}
wf2 <- workflow() %>%
  add_recipe(rec2)
```

# 6.6 Fitting the models

```{r}
log_fit2 <- wf2 %>% add_model(log_spec2) %>% fit(train2)
rforest_fit2  <- wf2 %>% add_model(rforest_spec2) %>% fit(train2)
xgboost_fit2 <- wf2 %>% add_model(xgboost_spec2) %>% fit(train2)

```

# 6.7 Evaluating the models

```{r}

# LOGISTIC
log_pred2 <- predict(log_fit2, test2, type = "prob") %>%
  bind_cols(test2) %>%
  mutate(pred_class = as.factor(ifelse(.pred_1 >= 0.5, 1, 0)))

log_conf2 <- log_pred2 %>%
  conf_mat(truth = is_misinformation, estimate = pred_class)

log_acc2 <- log_pred2 %>%
  accuracy(truth = is_misinformation, estimate = pred_class)

log_conf2 # view in the terminal 
log_acc2
# 0.61

# RFOREST
pred_rforest2 <- predict(rforest_fit2, test2) %>%
  bind_cols(test2)

conf_mat(pred_rforest2, truth = is_misinformation, estimate = .pred_class)
accuracy(pred_rforest2, truth = is_misinformation, estimate = .pred_class)
# 0.58


#XBOOST
pred_xgboost2 <- predict(xgboost_fit2, test2) %>%
  bind_cols(test2)

conf_mat(pred_xgboost2, truth = is_misinformation, estimate = .pred_class)
accuracy(pred_xgboost2, truth = is_misinformation, estimate = .pred_class)
# 0.57


```

# Confusion matrices

```{r}
# for log
cm_log_tbl <- as.table(log_conf$table) %>%
  as_tibble() %>%
  rename(prediction = Prediction, truth = Truth, n = n) %>%
  mutate(truth = as.factor(truth), prediction = as.factor(prediction))

ggplot(cm_log_tbl, aes(x = truth, y = prediction, fill = n)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = n), color = "white", fontface = "bold", size = 6) +
  scale_fill_gradient(low = "#d4e6f1", high = "#1f618d") +
  labs(title = "Confusion matrix – Logistic regression", subtitle = "Without penalty", x = "True classification", y = "Predicted classification") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5), plot.subtitle = element_text(size = 12, hjust = 0.5), panel.grid = element_blank()) +
  coord_fixed()


```

```{r}
# for Rforest
cm_rforest_tbl <- as.table(conf_mat(pred_rforest2, truth = is_misinformation, estimate = .pred_class)$table) %>%
  as_tibble() %>%
  rename(prediction = Prediction, truth = Truth, n = n) %>%
  mutate(truth = as.factor(truth), prediction = as.factor(prediction))

ggplot(cm_rforest_tbl, aes(x = truth, y = prediction, fill = n)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = n), color = "white", fontface = "bold", size = 6) +
  scale_fill_gradient(low = "#d4e6f1", high = "#1f618d") +
  labs(title = "Confusion matrix – Rforest", x = "True classification", y = "Predicted classification") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5), panel.grid = element_blank()) +
  coord_fixed()

```

```{r}
# for XGBoost
cm_xgb_tbl <- as.table(conf_mat(pred_xgboost2, truth = is_misinformation, estimate = .pred_class)$table) %>%
  as_tibble() %>%
  rename(prediction = Prediction, truth = Truth, n = n) %>%
  mutate(truth = as.factor(truth), prediction = as.factor(prediction))

ggplot(cm_xgb_tbl, aes(x = truth, y = prediction, fill = n)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = n), color = "white", fontface = "bold", size = 6) +
  scale_fill_gradient(low = "#d4e6f1", high = "#1f618d") +
  labs(title = "Confusion matrix – XGBoost", x = "True classification", y = "Predicted classification") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5), panel.grid = element_blank()) +
  coord_fixed()
```

# 6.8 Further evaluation of the models

```{r}

# introduce the evaluation metrics 
misinfo_metrics <- metric_set(accuracy, precision, recall, f_meas) #f-score = f_meas

# COMPUTE METRICS FOR EACH MODEL 

# (1) log
metrics_log <- misinfo_metrics(
  data = log_pred2,
  truth = is_misinformation,
  estimate = pred_class,
  event_level = "second"    
) |> 
  mutate(model = "Logistic regression")

# (2) rforest
metrics_rf <- misinfo_metrics(
  data = pred_rforest2,
  truth = is_misinformation,
  estimate = .pred_class,
  event_level = "second"
) |> 
  mutate(model = "Random forest")

#3 XGBoost
metrics_xgb <- misinfo_metrics(
  data = pred_xgboost2,
  truth = is_misinformation,
  estimate = .pred_class,
  event_level = "second"
) |> 
  mutate(model = "XGBoost")

# Table-view
all_metrics <- bind_rows(metrics_log, metrics_rf, metrics_xgb)

all_metrics
```

# visualize the metrics

```{r}
metrics_plot <- all_metrics |>
  mutate(metric = recode(.metric,accuracy  = "Accuracy", precision = "Precision", recall = "Recall", f_meas = "F1-score")) |>
  ggplot(aes(x = metric, y = .estimate, fill = model)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(title = "Performance by model", x = NULL, y = "Score", fill = "Model") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

metrics_plot

```

# STEP 7 - Tuning 2 - introducing penalties

# 7.7 NEW - Evaluating the models

```{r}
# introducing a threshold that aims to penalize false positives (truthful signal when misinformation) more than false negatives (misinformation signal when truthful)
threshold <- 0.45

log_pred2_with_threshold <- predict(log_fit2, test2, type = "prob") %>%
  bind_cols(test2) %>%
  mutate(pred_class = as.factor(ifelse(.pred_1 >= threshold, 1, 0))) # adjusted

log_conf2_with_threshold <- log_pred2_with_threshold %>%
  conf_mat(truth = is_misinformation, estimate = pred_class)

log_acc2_with_threshold <- log_pred2_with_threshold %>%
  accuracy(truth = is_misinformation, estimate = pred_class)

log_conf2_with_threshold
log_acc2_with_threshold

# without threshold => 0.61
# threshold = 0.4 => 0.55
# threshold = 0.45 => 0.56
```

```{r}
# RFOREST + probabilities
rforest_prob2 <- predict(rforest_fit2, test2, type = "prob") %>%
  bind_cols(test2)

# applying threshold
rforest_pred_threshold <- rforest_prob2 %>%
  mutate(pred_class_threshold = factor(ifelse(.pred_1 >= threshold, "1", "0"), levels = c("0", "1")))

# confusion matrix
rforest_cm_threshold <- rforest_pred_threshold %>%
  conf_mat(truth = is_misinformation, estimate = pred_class_threshold)

rforest_cm_threshold

```

# 7.8

```{r}
#XBOOST + probabilities
xgboost_prob2 <- predict(xgboost_fit2, test2, type = "prob") %>%
  bind_cols(test2)

# + threshold
xgboost_pred_threshold <- xgboost_prob2 %>%
  mutate(pred_class_threshold = factor(ifelse(.pred_1 >= threshold, "1", "0"), levels = c("0", "1")))

# confusion matrix
xgboost_conf_threshold <- xgboost_pred_threshold %>%
  conf_mat(truth = is_misinformation, estimate = pred_class_threshold)

xgboost_conf_threshold
```

# confusion matrices

```{r}
# log + threshold
cm_log_tbl <- as.table(log_conf2_with_threshold$table) %>%
  as_tibble() %>%
  rename(prediction = Prediction, truth = Truth, n = n) %>%
  mutate(truth = as.factor(truth), prediction = as.factor(prediction))

ggplot(cm_log_tbl, aes(x = truth, y = prediction, fill = n)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = n), color = "white", fontface = "bold", size = 6) +
  scale_fill_gradient(low = "#d4e6f1", high = "#1f618d") +
  labs(title = "Confusion matrix – Logistic regression*", subtitle = "*With penalty (0.45)", x = "True classification", y = "Predicted classification") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5), plot.subtitle = element_text(size = 12, hjust = 0.5), panel.grid = element_blank()) +
  coord_fixed()

```

![]()

```{r}
# Rforest + threshold
cm_rforest_with_threshold_tbl <- as.table(rforest_cm_threshold$table) %>%
  as_tibble() %>%
  rename(prediction = Prediction, truth = Truth, n = n) %>%
  mutate(truth = as.factor(truth), prediction = as.factor(prediction))

ggplot(cm_rforest_with_threshold_tbl, aes(x = truth, y = prediction, fill = n)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = n), color = "white", fontface = "bold", size = 6) +
  scale_fill_gradient(low = "#d4e6f1", high = "#1f618d") +
  labs(title = "Confusion matrix – Rforest*", subtitle = "*With penalty (0.45)", x = "True classification", y = "Predicted classification") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5), plot.subtitle = element_text(size = 12, hjust = 0.5), panel.grid = element_blank()) +
  coord_fixed()

```

```{r}
#  XGBoost + penalty
cm_xgboost_with_threshold_tbl <- as.table(xgboost_conf_threshold$table) %>%
  as_tibble() %>%
  rename(prediction = Prediction, truth = Truth, n = n) %>%
  mutate(truth = as.factor(truth), prediction = as.factor(prediction))

ggplot(cm_xgboost_with_threshold_tbl, aes(x = truth, y = prediction, fill = n)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = n), color = "white", fontface = "bold", size = 6) +
  scale_fill_gradient(low = "#d4e6f1", high = "#1f618d") +
  labs(title = "Confusion matrix – XGBoost*", subtitle = "*With penalty (0.45)", x = "True classification", y = "Predicted classification") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5), plot.subtitle = element_text(size = 12, hjust = 0.5), panel.grid = element_blank()) +
  coord_fixed()
```

# adjusting other metrics

```{r}
# log + threshold
metrics_log <- misinfo_metrics(
  data = log_pred2_with_threshold,
  truth = is_misinformation,
  estimate = pred_class,
  event_level = "second"    
) |>
  mutate(model = "Logistic regression")

# random forest + threshold
metrics_rf <- misinfo_metrics(
  data = rforest_pred_threshold,
  truth = is_misinformation,
  estimate = pred_class_threshold,  
  event_level = "second"
) |>
  mutate(model = "Random forest")

# XGBoost + threshold
metrics_xgb <- misinfo_metrics(
  data = xgboost_pred_threshold,
  truth = is_misinformation,
  estimate = pred_class_threshold,   
  event_level = "second"
) |>
  mutate(model = "XGBoost")

# combine all in one table
all_metrics <- bind_rows(metrics_log, metrics_rf, metrics_xgb)

all_metrics
```

# visualize the metrics

```{r}
metrics_plot <- all_metrics |>
  mutate(metric = recode(.metric,accuracy  = "Accuracy", precision = "Precision", recall = "Recall", f_meas = "F1-score")) |>
  ggplot(aes(x = metric, y = .estimate, fill = model)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(title = "Performance by model*", subtitle = "*With penalty (0.45)", x = NULL, y = "Score", fill = "Model") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5), plot.subtitle = element_text(size = 12, hjust = 0.5))

metrics_plot
```
